---
title: "Beak"
author: "Karim Dahr, Brendan Fox, Tingwei Hu, Joseph Milosh"
date: "2022-12-09"
output: html_document
bibiography: Beakering.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(knitr)
library(Momocs)
library(scatterplot3d)
library(viridis)
library(RColorBrewer)
library(reshape2)
```
# Introduction




# Methods
```{r, data prep, message=FALSE}
data.ori <- read_csv("AVONETdata_BirdTree_Pigot2020.csv")

#removed any rows where Trophic.Level or Trophic.Niche has N/A value
data.tmp <- data.ori[!is.na(data.ori$Trophic.Level),]
data <- data.tmp[!is.na(data.tmp$Trophic.Niche),]
```

```{r, normalized ratios & log transform relevant columns}
data.upd <- data %>% 
  mutate(l.w.ratio = Beak.Length_Culmen / Beak.Width) %>% 
  mutate(bsize.mass.ratio = l.w.ratio / Mass) %>% #beak lw ratio/body mass
  mutate(l.log = log10(data$Beak.Length_Culmen)) %>% 
  mutate(w.log = log10(data$Beak.Width)) %>% 
  mutate(mass.log = log10(data$Mass))
```

## Data Collection
Morphological and ecological data analyzed in this report were gathered from the AVONET data set described by @tobias2022avonet. Within the data set, geographical and phylogenetic information were organized and the results shared by @pigot2020macroevolutionary. 
## Data Analysis


# Results



# Discussion


# Author Contributions
Karim Dahr: 
Brendan Fox:
Tingwei Hu:
Joseph Milosh: 

# References